Ownership types in theory and practice (in Rust)

### 2023-01-22 slides/slides.tex
finished transferring ownership slides

\begin{frame}{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extended version of owners-as-modifiers
  \begin{itemize}
    \item immutable (read-only) and mutable references allowed (not in owners-as-dominators)
    \item no garbage collection (overhead)
    \item extensions:
    \begin{itemize}
      \item transferring ownership from one object to another
      \item object immutability by default
    \end{itemize}
  \end{itemize}
  % \item aliasing concept essential to ownership types
  % \item different ways of implementing aliasing:
  % \begin{itemize}
  %   \item pointers: exist in Rust, but their use is not advised
  %   \item \emph{references}
  % \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{Transferring ownership}
  Big difference to theoretical ownership types
\end{frame}

@@

@@

% \subsection{Another ownership implementation}

>>>>

\begin{frame}{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extends owners-as-modifiers
  \begin{itemize}
    \item no garbage collection (overhead)
    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
    \item rules for compiler to check validity of references
    \item extensions:
    \begin{itemize}
      \item transferring ownership from one object to another
      \item object immutability by default
      \item Resource Acquisition Is Initialization (similar to C++)
    \end{itemize}
  \end{itemize}
  % \item aliasing concept essential to ownership types
  % \item different ways of implementing aliasing:
  % \begin{itemize}
  %   \item pointers: exist in Rust, but their use is not advised
  %   \item \emph{references}
  % \end{itemize}
  \item ownership rules in Rust:
  \begin{enumerate}
    \item Each value in Rust has an owner.
    \item There can only be one owner at a time.
    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
  \end{enumerate}
\end{itemize}
\end{frame}

@@

% TODO: variable scope

\begin{frame}[fragile]{Transferring ownership}
Heap-based values: 
\begin{itemize}
  \item unknown size at compile time
  \item copying the value at runtime could be costly
  \item values are "moved" (between variables) by default

  \begin{lstlisting}[language=Rust]
  fn main() {
    let a = String::from("hello world");
    let b = a; // ownership transfer: a is moved into b
    let c = take_ownership(b); // b is moved into function
    println!("{}", b); // compiler error: b no longer valid
  }
  fn take_ownership(s: String) -> String {
    // do something with s
    String::from("new hello world")
  }
  \end{lstlisting}
  \item create deep copies with explicit \verb|clone()| call
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Transferring ownership}
Stack-based primitives: 
\begin{itemize}
  \item small, known size at compile time
  \item examples: i32, u8, f64, char, bool etc.
  \item values are copied by default

  \begin{lstlisting}[language=Rust]
  let x: u32 = 42;
  let y: u32 = x; // x is copied instead of moved
  println!("x: {}, y: {}", x, y); // x and y valid
  \end{lstlisting}
  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
\end{itemize}
\end{frame}

@@

\begin{frame}{Lifetimes}
\end{frame}

\begin{frame}[fragile]{References and borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let a;
  {
    let b: u32 = 42;
    a = &b;
  }
  println!("{}", *a) // dereference a
  \end{lstlisting}

  \item Will it compile? \pause No.
  % TODO: lifetime explanation
  \item Why not? \pause Borrowed value 'b' does not live long enough/is used after being dropped.
  \item Refactored version that compiles:\pause

  \begin{lstlisting}[language=Rust]
  let a;
  let b: u32 = 42;
  a = &b;
  println!("{}", *a) // dereference a
  \end{lstlisting}
\end{itemize}
\end{frame}

@@
